Reasonably Programmable Literal Notation

### 2017-11-12 pldi18/syntax-pldi18.tex
work on intro

General-purpose programming languages typically define literal notation for only a small number of data structures, e.g. lists. This design practice is unsatisfying because there are many other data structures for which literal notation might be useful, e.g. finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

@@

This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles. We are integrating TLMs into Reason, an increasingly popular alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common to build in \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few standard data structures. For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This language design practice is decidedly \emph{ad hoc} -- it is quite easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. For example, 1) clients of a ``collections'' library might also want matrix, set and map literals following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although requests for new literal notations are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, file system paths, URLs and many other data structures. Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

@@

\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e. by inspecting only the surface syntax, about such fundamental issues as types and variable binding when a program text is littered with  unfamiliar user-defined literal forms. Instead, programmers can only reason transparently, i.e. by inspecting the underlying expansion or the extension responsible for producing the expansion. In some existing systems, even determining which extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''

\begin{figure}[t!]
\vspace{-2px}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-14px}
\label{fig:K-dialect}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason} that has, hypothetically, been extended with some large number of new literal forms by a grammar extension system, like Camlp4 \cite{ocaml-manual} or Sugar* \cite{erdweg2011sugarj,erdweg2013framework}. 
The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot fall back on any abstract reasoning principles to answer questions like:

\begin{enumerate}[nolistsep,noitemsep,leftmargin=10pt]
\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3? Might different extensions generate different expansions for this form (i.e. is there syntactic ambiguity)?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

@@

\paragraph{Approach}

This paper formally defines a more reasonable mechanism for defining new literal notation: \emph{typed literal macros (TLMs)}. TLMs give library providers  programmatic control over the parsing and expansion of \emph{generalized literal forms} \cite{TSLs}, which syntactically subsume other literal forms because their bodies are left initially unparsed according to the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.)

\begin{figure}
\vspace{-2px}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
\vspace{-14px}
\label{fig:K-tsm-example}
\end{figure}

For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example from Figure \ref{fig:K-dialect} just discussed. The applied TLM parses the literal body (i.e. the characters between the delimiters) and generates an expansion statically. During the typing phase, a validation step ensures that programmers can reason abstractly about the five issues just identified. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly about Line 3 as follows:
\begin{enumerate}[noitemsep,nolistsep,leftmargin=10pt]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax that this parser implements is unambiguous (even in situations where a spliced term itself applies a TLM, not shown here.)
\item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} A type annotation on the definition of \li{$kquery} (not shown) determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\vspace{-4px}
\subsection{Splicing}\label{sec:splicing-and-hygiene}
\vspace{-2px}
When the parse function determines that some segment of the literal body should be taken as a spliced expression, according to whatever syntactic criteria it deems suitable, it must not directly insert the syntax tree of that expression into the encoding of the expansion. Instead, 
the TLM must indirectly refer to the spliced expression using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body. The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have. Types can be spliced out by using the \li{SplicedT} variant of \li{proto_typ} analagously.

@@

For example, consider again the TLM application on Line 1 of Figure \ref{fig:first-tsm-example}:
\li{$html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]}. The proto-expansion generated by \li{$html}, if written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, is:
\begin{lstlisting}[numbers=none]
  H1Element(Nil, Cons(TextNode "Hello, ", Cons(
                      TextNode spliced<13; 22; string>, Nil)))
\end{lstlisting}
The proto-expression \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded. It corresponds to the following encoding:
\li{SplicedE (\{startIdx: 13, endIdx: 22\}, StringTy)}.

@@

% \subsection{Splice Summaries and Segmentations}
The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within. For example, the segmentation inferred from the proto-expression above is the finite set containing only \li{spliced<13; 22; string>}. % Notice that no information about  the spliced terms appear is communicated by the splice summary.
The semantics checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
This resolves the problem of \textbf{Segmentation} described in Sec. \ref{sec:intro}, i.e. every literal body in a well-typed program has a well-defined segmentation. A program editor or pretty-printer can communicate this location information to the programmer, e.g. by coloring non-spliced segments green as is our convention in this document. In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth. For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading ($str [|SSTRWorld!ESTR|])SURL\} ...</body>EURL|]}.

>>>>

General-purpose programming languages typically define literal notation for only a small number of data structures, e.g. lists. This is unsatisfying because there are many other data structures for which literal notation might be useful, e.g. finite maps, regexes, HTML data, SQL queries, syntax trees and encodings of chemical structures. %A less \emph{ad hoc} design would be one that allows library providers to define new literal forms in a decentralized manner. However, the mechanisms available to programmers today, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are \emph{unreasonable}. In particular, they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.)% Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, defeating much of the purpose of convenient literal forms. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

@@

This paper establishes these abstract reasoning principles formally with a calculus of typed expressions, patterns and modules. 
%This calculus is the first detailed type-theoretic account of a hygienic macro system, of any design, for a language with these essential features of ML, and the first macro system with support for literal parsing that comes equipped with these essential reasoning principles.
We are integrating TLMs into Reason, an alternative front-end for OCaml.% This calculus is the first such formalization of a typed macro system.

@@

% \vspace{-4px}
\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
% The surface syntax of a programming language is its user interface. 
When designing the surface syntax of a general-purpose programming language, it is common practice to define \emph{literal notation} that decreases the syntactic cost of constructing and pattern matching over values of a few data structures. For example, in Standard ML (SML), the list literal \li{[x, y, z]} is equivalent by desugaring to \li{Cons(x,Cons(y,Cons(z,Nil)))} where \li{Nil} and \li{Cons} stand for the constructors of the standard \li{list} datatype \cite{mthm97-for-dart,harper1997programming}. 
%\footnote{In SML, list literals work even in contexts where the list expression constructors have been shadowed by other bindings, i.e. the meaning of literal forms is \emph{context independent}. 
% We will return to this concept again later.} 
Similarly, Ur/Web's surface syntax extends Ur's syntax with expression and pattern literals for encodings of XML and HTML data \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} shows two such HTML literals, one that ``splices in'' a string expression (Line 1) and the other an HTML expression (Line 2).
%The desugarings of these literal expressions, not shown, are substantially more verbose and less readable to programmers familiar with XHTML.

This language design practice is decidedly \emph{ad hoc}, in that it is easy to come up with other examples of data structures that are not privileged with literal forms in these languages, but for which mathematicians, scientists or programmers have invented specialized notation \cite{DBLP:journals/cacm/Iverson80,cajori1928history}. For example, 1) clients of a ``collections'' library might also want matrix, set and map literals following usual mathematical conventions; 2) clients of a ``web programming'' library might like not just HTML literals but also CSS literals (which Ur/Web lacks); 3) a compiler writer might like ``quotation'' literals for the terms of the object language and various intermediate languages; and 4) clients of a ``chemistry'' library might like chemical structure literals based on the SMILES standard \cite{anderson1987smiles}.

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the prevalence of string injection attacks, which are amongst the most common and serious security vulnerabilities on web servers today \cite{owasp2017}. More generally, evidence suggests that programmers frequently engage in ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that about 15\% of them were parseable according to some readily apparent type-specific grammar, e.g. for SQL queries, regexes, URLs and many other data structures. Literal forms, with support for splicing, would decrease the syntactic cost of composite encodings of such data, which are more amenable to programmatic manipulation and compositional reasoning than flat string encodings. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

@@

\paragraph{Problem} The problem that motivates this paper is that it is impossible to reason abstractly, i.e. by inspecting only the surface syntax, about such fundamental issues as types and variable binding when the program text is littered with  unfamiliar user-defined literal forms. Instead, programmers can only reason transparently, i.e. by inspecting the underlying expansion or the extension responsible for producing the expansion. In some existing systems, even determining which extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs (modules), possibly written by different people''

\begin{figure}[t!]
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = [|(!R)@&{&/x!/:2_!x}'!R}|];
\end{lstlisting}
\vspace{-10px}
\caption{Syntax extension systems produce unreasonable program text.}
\vspace{-11px}
\label{fig:K-dialect}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend, i.e. reason about, the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's syntax called Reason \cite{reason} that has, hypothetically, been extended with some large number of new literal forms by a grammar extension system, like Camlp4 \cite{ocaml-manual} or Sugar* \cite{erdweg2011sugarj,erdweg2013framework}. 
The literal form on Line 3 constructs an encoding of a query in the niche stack-based database query language K, using its intentionally terse notation \cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer unfamiliar with this notation cannot fall back on abstract reasoning principles to answer questions like:

\begin{enumerate}[leftmargin=10pt]
\item \textbf{(Responsibility)} Which syntax extension determined the expansion of the literal form on Line 3? Might different extensions generate different expansions for this form (i.e. is there syntactic ambiguity)?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions (meaning that they appear directly in the underlying expansion), or are they parsed in some other way peculiar to this literal form (e.g. as operators in the K query language)?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2, or might it capture an unseen binding of the same identifier in the expansion of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Lines 1-2, is renamed, could that possibly break Line 3 or change its meaning because the expansion assumes that \li{w} is in scope?% What other hidden assumptions might the desugaring be making?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

@@

An approach that has made some headway against these problems is Omar et al.'s \emph{type-specific languages} (TSLs) \cite{TSLs}. In this approach, the language delegates control over the parsing and expansion of certain \emph{generalized literal forms} to a parse function associated with the type that the form is being checked against. Generalized literal forms syntactically subsume other literal forms because their bodies (i.e. the characters between the outer delimiters, here \li{[|} and \li{|]}) are left  unparsed by the fixed context-free syntax of the language (i.e. they are initially parsed like raw string literals.) With TSLs, Line 3 of our example would require a type annotation:
\begin{lstlisting}[numbers=none]
    let y : kquery = [|(!R)@&{&/x!/:2_!x}'!R}|]
\end{lstlisting}
The parse function associated with the \li{kquery} type would be invoked to parse and expand the literal body.

Although TSLs address the problems of \textbf{Responsibility} and \textbf{Typing}, there is no guarantee that a valid segmentation exists, nor was the problem of reasoning about binding (i.e. capture and context dependence) adequately addressed in the prior work. Moreover, this prior work was introduced in the context of a very simple monomorphic, nominally typed language. There was no support for important features of modern languages in the ML tradition, such as structural types (e.g. arrow and tuple types), parameterized types, abstract types, modules, pattern notation and ML-style type inference.  Finally, parse functions and the expansions they generate needed to be closed, i.e. they could not make use of any libraries at all. Taken together, we found that TSLs were not suitable as-is for integration into Reason, and other languages with a broadly similar type and binding discipline (e.g. Haskell, Scala, Rust and so on.)

\paragraph{Contributions} The aim of this paper is to address all of these deficiencies, resulting in the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) that is expressive enough for integration into a full-scale typed functional language (in this paper, the Reason front-end for OCaml.) We evaluate the first claim by developing a series of formal calculi equipped with proofs of these reasoning principles, and the second claim with several non-trivial case studies that appear throughout the paper. The empirical evaluation in the prior work on TSLs \cite{TSLs} supports the claim that a mechanism for defining new literal notation is useful, so our focus is on formal reasoning principles and integration with advanced language features in this paper.

\begin{figure}
\begin{lstlisting}
  let w = compute_w();
  let x = compute_x(w);
  let y = $kquery [|SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL|];
\end{lstlisting}
\vspace{-10px}
\caption{TLMs make examples like the one from Figure \ref{fig:K-dialect} more reasonable.}
\vspace{-10px}
\label{fig:K-tsm-example}
\end{figure}

The most obvious point of departure between TSLs and \emph{typed literal macros} (TLMs), which are the topic of this paper, is that we separate the mechanism for defining new literal notation from the type definition mechanism of the language. TLMs are explicitly named and applied. For example, Figure \ref{fig:K-tsm-example} applies a TLM named \li{$kquery} to a generalized literal form delimited by \li{[|} and \li{|]} to express the K query example just discussed. The applied TLM parses the literal body and generates an expansion statically.

During the typing phase, a validation step ensures that programmers can reason abstractly about the five issues just identified. We will characterize these reasoning principles precisely as we proceed, but for now, let us develop the basic intuition -- the programmer can reason abstractly as follows:
\begin{enumerate}[nolistsep,leftmargin=10pt]
\item \textbf{(Responsibility)} The applied TLM, \li{$kquery}, parses and determines the expansion of the literal body. The context-free syntax of the language is fixed, so the TLM provider is able to modularly establish that the syntax that this parser implements is unambiguous (even in situations where a spliced term itself applies a TLM, not shown here.)
\item \textbf{(Segmentation)} The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms from segments that are parsed in some other way by the TLM. The segmentation can be communicated by a code editor using secondary notation, e.g. colors in this document. So by examining Figure \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators in the K language (because they are  green.)
\item \textbf{(Capture)} Splicing is capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot capture a coincidental binding of \li{x} in the expansion.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, e.g., \li{w} is in scope. The client is free to rename \li{w}.
\item \textbf{(Typing)} A type annotation on the definition of \li{$kquery} (not shown) determines the type that the expansion must have. We will see an example of a TLM definition in the next section. Moreover, each spliced segment in the segmentation also comes with a type annotation. %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

@@

\vspace{-4px}
\subsection{Splicing}\label{sec:splicing-and-hygiene}
\vspace{-2px}
When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it using the \li{SplicedE} variant of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body. The \li{SplicedE} variant also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have. Types can be spliced out by using the \li{SplicedT} variant of \li{proto_typ} analagously.

@@

For example, consider again the TLM application on Line 1 of Figure \ref{fig:first-tsm-example}:
\li{$html [|SURL<h1>Hello, {[EURLfirst_nameSURL]}</h1>EURL|]}. The parse function of the \li{$html} TLM (Figure \ref{fig:html-tlm-def}, Lines 2-4) first sends the literal body through a pre-existing HTML parser, \li{parse_html}\todo{which one?}. It then passes the result to a function \li{html_to_ast}, not shown, which produces a value of type \li{proto_expr}. When this function encounters a text node containing matched \li{\{[} and \li{]\}}, then that segment is inserted as a spliced expression of type \li{string}. Similarly, text nodes containing matched curly braces produce a spliced expression of type \li{html}. We assume that the pre-existing HTML parser annotates each element or node with its position in the originally parsed text (as most production-grade HTML/XML parsers are capable of doing.) If written in a textual syntax for proto-expressions where references to spliced expressions are written \li{spliced<startIdx; endIndex; ty>}, the output of this function for the example above is:
\begin{lstlisting}[numbers=none]
    H1Element(Nil, Cons(TextNode "Hello, ", Cons(
      TextNode spliced<13; 22; string>, Nil)))
\end{lstlisting}
The proto-expression \li{spliced<13; 22; string>} is a reference to the spliced string expression \li{first_name} by its location relative to the start of the literal body being expanded. It corresponds to the following encoding:
\begin{lstlisting}[numbers=none]
    SplicedE ({startIdx=13, endIdx=22}, StringTy)
\end{lstlisting}

@@

% \subsection{Splice Summaries and Segmentations}
The \emph{segmentation} inferred from a proto-expansion is the finite set of references to spliced terms contained within. For example, the segmentation inferred from the proto-expression above contains only \li{spliced<13; 22; string>}. % Notice that no information about  the spliced terms appear is communicated by the splice summary.
The system checks that all of the locations in the segmentation are 1) in bounds relative to the literal body; and 2) non-overlapping. 
This resolves the problem of \textbf{Segmentation} described in Sec. \ref{sec:intro}, i.e. every literal body in a well-typed program has a well-defined segmentation. A program editor or pretty-printer can communicate this location information to the programmer, e.g. by coloring non-spliced segments green as is our convention in this document. In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth. For example, if strings were not primitive but rather defined as sequences of characters, we might define a TLM \li{$str} to recover string literal notation and write \li{$html [|SURL<body>\{EURLheading $str [|SSTRWorld!ESTR|]SURL\} ...</body>EURL|]}.
